---
title: "EDA of Prosper Loan Data by Vaikunth Kannan"
output: html_document
---

#Abstract
Prosper.com is a P2P lending marketplace, with more than 2 million members. Udacity last updated this data on 03/11/2014.

```{r packages, cache=TRUE, echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(gmodels)
library(knitr)
library(lubridate)
library(RColorBrewer)
library(GGally)
library(reshape)
library(gridExtra)
library(tidyr)
library(MASS)
library(scales)
```

#Analysis

#Overview of given dataset
```{r echo=FALSE, Load_the_Data}
setwd('/eda')
#Read CSV file
prosper = read.csv("prosperLoanData.csv")
```

#Browse the data
This dataset contains 113937 observations with 81 variables.

Here is the structure of the dataset
```{r echo= FALSE, structure}
names(prosper)
str(prosper)
dim(prosper)
```

#The main variables which influence the dataset are the borrower characteristics and the propser rating given to each of them.
```{r echo= FALSE, data_clean}
prosper$IncomeRange <- factor(prosper$IncomeRange, 
                           levels=c("Not employed", "$0", "$1-24,999", 
                                    "$25,000-49,999", "$50,000-74,999", 
                                    "$75,000-99,999", "$100,000+", 
                                    "Not displayed"))
prosper$ProsperRating.alpha = factor(prosper$ProsperRating..Alpha.,
                                 levels = c("AA","A","B","C","D","E","HR","NA"))
prosper$ProsperRating <-factor(prosper$ProsperRating..Alpha,
                        levels = c('AA', 'A', 'B', 'C', 'D', 'E', 'HR', 'NA'))
prosper$ProsperScore = factor(prosper$ProsperScore)
prosper$ListingCreationDate <- ymd_hms(prosper$ListingCreationDate)
prosper$IsBorrowerHomeowner <- factor(prosper$IsBorrowerHomeowner)
```

#Univariate Analysis
Annual income of borrowers
```{r echo= FALSE, annual_income}
qplot(data = prosper, x= IncomeRange, color = I('black'), fill = I('#068EDB')) +labs(title='The Income Range of the Borrowers')
```
It looks like a normal distribution with the income of most borrowers in the range of 25,000 - 74,999. There are also a significant amount of people with no income who got loans. We need to explore how this income was calculated and how they got loans.

#Histogram of prosper rating
```{r echo= FALSE, hist_prosper}
qplot(data = na.omit(prosper), x = ProsperRating.alpha, fill = I('orange')) +
  ggtitle('Numbers of Loans by Prosper Rating') +
  xlab('Rating') +
  ylab('Number of Loans')
summary(prosper$ProsperRating.alpha)
```
Loan rating with C are the majority and there are also high prosper rating borrowers.

#Prosper Score Distribution
```{r echo=FALSE, hist_score}
qplot(data = prosper, x = ProsperScore, color = I('black'), fill = I('#007EE5')) + 
        theme(axis.text.x = element_text(angle = 60, vjust = 0.6)) +
        xlab("Score") + ggtitle("Prosper Score Distribution")
summary(prosper$ProsperScore)
```
The rating lies in the 2 to 10 range with most of the loans not rated.

```{r echo=FALSE, Debt_to_Income_Ratio}
##4. Debt to Income Ratio
qplot(data = prosper, x = DebtToIncomeRatio, color = I('black'), fill = I('red'), binwidth = 0.1) +                
        xlim(0, quantile(prosper$DebtToIncomeRatio, prob = 0.99, na.rm=TRUE)) +
        ggtitle("Debt To Income Ratio") +
        xlab("Debt to Income Ratio") +
        ylab("Count")
summary(prosper$DebtToIncomeRatio)
```
The ratio is around 0.25 which is 25% for most of the borrowers which shows good credit history for the borrowers.

One thing I found on the website was that the terms of the loans were three years long. Lets verify it.
```{r prosper_term, cache=TRUE, echo=TRUE, echo=FALSE}
cat("Frequecy Table for Term of the loans")
CrossTable(prosper$Term) 
prosper$Term <- factor(prosper$Term) #For later Analysis
```
Indeed, 3 year term is the most popular choice.

Lets find out whats the purpose of the loan borrower.
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=8, Borrower_Purpose}

x <- c('Debt Consolidation','Home Improvement','Business', 
                                 'Personal Loan', 
                                 'Student Use', 
                                 'Auto', 
                                 'Baby & Adoption',
                                 'Boat', 
                                 'Cosmetic Procedure', 
                                 'Engagement Ring', 
                                 'Green Loans', 
                                 'Household Expenses', 
                                 'Large Purchases', 
                                 'Medical/Dental', 
                                 'Motorcycle', 'RV',
                                 'Taxes', 'Vacation',
                                 'Wedding Loans', 
                                 'Other', 
                                 'Not Available')
prosper$ListingCategory <- factor(prosper$ListingCategory..numeric., 
                                  levels = c(1:6,8:20,7,0), labels = x)

qplot(data = prosper, x = prosper$ListingCategory, color = I('black'), fill = I('blue')) +
                theme(axis.text.x = element_text(angle = 90, vjust = 0.6)) + 
                ggtitle("The Purpose of the Loan Listing") +
                xlab("Listing Category") +
                ylab("Number of loans")
summary(prosper$ListingCategory)
```
So many people give the purpose as debt consolidation. Also, a lot of loans have no purpose which needs to be investigated. As usual, significant number of people have invested in Auto, Household and Medical Expenses.

Lets take a look at the Loan Start date
```{r echo= FALSE, loan_date}
prosper$LoanOriginationDate <- as.POSIXct(prosper$LoanOriginationDate,
                                                    format="%Y-%m-%d")
prosper$LoanOriginationDate.year <-prosper$LoanOriginationDate %>% 
                                          format('%Y') %>% strtoi()
qplot(data = prosper, x = as.factor(LoanOriginationDate.year), color = I('black'), fill =I('blue'))   + ggtitle('Number of Loans by Year') +
    xlab('Year') +
    ylab('Number of Loans')

summary(prosper$LoanOriginationDate.year)
```

Since the data is upto 2014 we can see that lot of borrowers are after 2012 which is a period of recovery after the global recession.

Splitting the Loan data by the month of borrowing.
```{r echo=FALSE, loan_by_month}
months <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun',
            'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
prosper$LoanOriginationDate.month=format(prosper$LoanOriginationDate, "%b")

qplot(data = prosper, x =LoanOriginationDate.month, color = I('black'), fill =I('blue')) +
  facet_wrap(~LoanOriginationDate.year) +
  ggtitle('Number of Loans by Month') +
  xlab('Month') +
  ylab('Number of Loans')
```
As you can see, A lot of activity is again in 2012 and that too at the start of the year and end of the year, where things stabilize and people have a clear idea about their finances

The amount of loan each borrower has taken.
```{r echo=FALSE, Loan_by_Amount}
qplot(data = prosper, x = LoanOriginalAmount, binwidth = 1000, color = I('black'), fill=I('red')) +                 scale_x_continuous(limits = c(0,quantile(prosper$LoanOriginalAmount, 0.99, 
                na.rm = TRUE)), breaks = seq(0, quantile(prosper$LoanOriginalAmount, 0.99, 
                na.rm = TRUE), 2000))

summary(prosper$LoanOriginalAmount)
```
The loan amount seems to be in the 2000-6000 range for the majority and there are a few people who have obtained at higher amounts which depends on the purpose of loan borrower, usually being house mortgage.

To get more insight, we can see whether people were employed when they obtained the loan which would define the purpose of loan.

```{r echo=FALSE, Employment_status}
qplot(x = EmploymentStatus, data = na.omit(prosper)) + ggtitle("Loans by Employment Status")
summary(prosper$EmploymentStatus)
```
This data seems satisfying as most of the people are employed and in full time jobs.

There are some cases where people are referred to banks to get loans. Lets see how many people obtained loans through referrals.
```{r recommendations, echo=FALSE}
qplot(x=Recommendations, data=prosper, binwidth = 1, color = I('black'), fill = I('red')) + 
    ggtitle("Number of Recommendations of the Borrowers")

table(prosper$Recommendations != 0, 
      dnn = "If the borrower has one or 
      more recommendations")
```

Since the value of 0 is high, majority of them did not have a recommendation.

Then, what was the monthly income of the borrowers that made them eligible for loans,

```{r echo=FALSE, Stated_Monthly_Income}
qplot(x = StatedMonthlyIncome, data = prosper, binwidth = 1000, color = I('white')) +
                  scale_x_continuous(limits = c(0, quantile(prosper$StatedMonthlyIncome, 0.99,
                  na.rm = TRUE)),breaks = seq(0, quantile(prosper$StatedMonthlyIncome, 0.99, 
                  na.rm = TRUE), 1000))
summary(prosper$StatedMonthlyIncome)
```
This correlates with the loan amount being borrowed being around the 4000 range since the mean and median montly income are in the same range and people are sensible to take loan which they are able to repay.

Looking at the bank's or lenders perspective, lets see if lenders are benefitting from the loan
```{r LenderYield, echo=TRUE}
summary(prosper$LenderYield)
qplot(x=LenderYield, data=prosper, binwidth = 0.005, color = I('black'), fill = I('red')) + 
    scale_x_continuous(breaks=seq(0, 0.5, 0.05)) + 
    labs(title="The Lender Yield on the Loan")

```

The plot shows that the lender yield is high at 0.31 which shows that lenders also had a good investment out of giving loans.

The number of investors investing in a loan can be determined to see how the funding occurs
```{r echo=FALSE, Total_Investor}
qplot(x=Investors, data= prosper, binwidth = 5, color = I('black'), fill = I('red')) + 
    ggtitle("Number of Investors Funded the Loan")

table(prosper$Investors==1)
```
Most loans are funded individually since the majority of the loan amounts are very small.

##Bivariate Analysis

Since the loan amount are pretty low, usually the loans could be from relatives or friends.
```{r InvestmentFromFriendsCount, echo=FALSE}
with(subset(prosper, Investors==1), table(InvestmentFromFriendsCount, 
           dnn = "if the single investor is 
           a friend of the borrower"))
```
But, from this data, borrowers have taken loans mostly not through friends.

Does the prosper rating mean there are lot of borrowers?
```{r echo=FALSE, Prospr_Rating}
prosper$ProsperRating.alpha <- factor(prosper$ProsperRating.alpha)
ggplot(data = prosper, aes(x = ProsperRating.alpha, y = BorrowerRate)) +
        geom_boxplot() +
        xlab("Prosper Rating") +
        ggtitle("Borrower Rate for Different Prosper Rating")
```
So, borrowers with higher prosper rating are not the majority ones to take loans.

Does the rating exactly describe the credibility and repayment capacity of the borrower?
```{r echo=FALSE, Defaulter}
prosper <- prosper %>% mutate(Status = ifelse(LoanStatus %in%
                     c("Chargedoff", "Defaulted"), 0,
                     ifelse(LoanStatus %in% 
                     c("Completed", "Current", "FinalPaymentInProgress"), 2, 
                     ifelse(LoanStatus %in% 
                     "Cancelled",3,1))))

prosper$Status <- factor(prosper$Status, levels = 0:3, 
                         labels = c("Defaulted", 
                                    "Past Due", 
                                    "Current or Paid",
                                    "Cancelled"))

ggplot(data = arrange(prosper,Status), aes(x = ProsperRating.alpha, 
                    y = LoanOriginalAmount, fill = Status)) +
                    geom_bar(stat = "identity") +
                    xlab("Prosper Rating") +
                    xlab("Original Loan Amount") +
                    ggtitle("Orignal Loan Amount for Different Prosper Rating")
```

The loans with E and HR have less defaulting but the number of loans in that category are pretty less. The higher the rating the number of loans defaulted has decreased adding to the authenticity of the rating.

Some people take the risk of obtaining loan with repayment higher than their monthly income. Lets see how many people have taken that risk to get a loan from prosper.
```{r income vs ratio, echo=TRUE}
ggplot(aes(x=DebtToIncomeRatio), data=prosper) +
    geom_density(aes(color = IncomeRange, 
                     fill  = IncomeRange), 
                 alpha = 0.5) +
    facet_wrap(~IncomeRange, ncol=2, 
               scales = "free")

with(prosper, by(DebtToIncomeRatio, IncomeRange, 
              summary))
```
People are putting a large portion of (even an amount exceeds) one's income every month to pay back the loan for 1, 3 or even 5 years continuously is extremely hard, if not a mission imporssible. But for borrowers who were not employed when they created the listings, more than a quarter of those people decided to pay the amount of money that is ten times of their monthly income for the loan.

Does the lender lending to borrowers with higher rating profit from the investment
```{r echo=FALSE, lendervsrating}
prosper$ProsperRating.alpha = factor(prosper$ProsperRating..Alpha.,
                                 levels = c("AA","A","B","C","D","E","HR",""))
ggplot(data = na.omit(prosper), aes(x = ProsperRating.alpha, y = LenderYield)) +
        geom_boxplot() + xlab("Prosper Rating") +
          ggtitle("Yield for Different Prosper Rating")
```

There seems to be an inverse relationship here.

The loan amount in terms of the repayment terms can be determined to get an idea if people are aware of the ability to repay higher loans.
```{r echo=FALSE, loanvsterm}
ggplot(aes(y = LoanOriginalAmount, x = Term), data = prosper) +
                                            geom_boxplot() +
                                            scale_y_continuous(
    limits = c(0, quantile(prosper$LoanOriginalAmount, 0.99, na.rm = TRUE)))
```
There is a direct relationship between the loan term and the amount of loan which is quite sensible and good.

##Multivariate Analysis

The lender yield should be dependent on the prosper rating which in turn should be dependent on the debt to income ratio.
```{r echo=FALSE, debtvslendervsrating}
ggplot(aes(x= DebtToIncomeRatio, y=LenderYield, color=ProsperRating.alpha), 
    data=na.omit(filter(prosper, DebtToIncomeRatio < 1))) +
    geom_point(alpha = 1) +
    #scale_y_log10() +
    #facet_grid(.~ ProsperRating.alpha ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    ggtitle("Lender Yield  vs Debt to Income Ratio vs Prosper Rate (for loans with rating") +
    xlab ("Debt To Income Ratio < 1") +
    ylab ("Lender Yield") +
    scale_fill_discrete(name = "Prosper Rating")
```

We can see that people are medium rating have higher lender yield and also less debt to income ratio.

Lets take a look if the year of the loan depends on the prosper rating.
```{r echo=FALSE, fig.width = 12, fig.height = 10, ratevsyearvsterm}

plot1 <- ggplot(aes(x= ProsperRating.alpha, y= BorrowerRate), 
       data=na.omit(filter(prosper))) +
    geom_boxplot(aes(fill = ProsperRating.alpha)) +
    facet_grid( .~ LoanOriginationDate.year) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    ggtitle("Lender Yield vs Term and Prosper Rate")

plot2 <- ggplot(aes(x= ProsperRating.alpha, y= BorrowerRate, color=ProsperRating.alpha ), 
    data=na.omit(filter(prosper))) +
    geom_point(alpha = 1, position = "jitter") +
    facet_grid( .~ LoanOriginationDate.year) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))+
    ggtitle("Lender Yield vs Term and Prosper Rate")

grid.arrange(plot1, plot2, ncol=1, nrow =2)
```
We can see that the borrower rates based on prosper rating shows stability over all the years establishing the fact that the data from propser is credible and the borrower pattern has not changed.

##Final Plots and Summary

###Plot one
The relationship between lender yield and prosper rating shows that the higher the risk, the lower the rating and better lender yield. Also people with higher rating have less debt to income ratio thus contributing to the prosper rating.
```{r echo=FALSE, fig.width = 12, fig.height = 8, Plot_One}
ggplot(aes(x= DebtToIncomeRatio, y=LenderYield, color=ProsperRating.alpha), 
       data=na.omit(filter(prosper, DebtToIncomeRatio < 1))) +
  geom_point(alpha = 1) +
  #scale_y_log10() +
  #facet_grid(.~ ProsperRating.alpha ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  ggtitle("Lender Yield  vs Debt to Income Ratio vs Prosper Rating") +
  xlab ("Debt To Income Ratio < 1") +
  ylab ("Lender Yield [%]") +
  scale_fill_discrete(name = "Prosper Rating")
```

The lender yield vs prosper rating shows that investors are willing to take risks by giving loans to people with less rating and defaulters. This describes the mentality of the people who do anything to earn profits and get more money
```{r echo=FALSE, fig.width = 12, fig.height = 8,Plot_Two}
ggplot(data = arrange(prosper,Status), aes(x = ProsperRating.alpha, 
                    y = LoanOriginalAmount, fill = Status)) +
                    geom_bar(stat = "identity") +
                    xlab("Prosper Rating") +
                    ylab("Original Loan Amount [$]") +
                    ggtitle("Orignal Loan Amount for Different Prosper Rating")
```
The final one is the loans taken by years which clearly depicts the recession period in USA and the confidence in people during those times to take loans.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width = 12, fig.height = 8, Plot_Three}
qplot(data = prosper,x = as.factor(LoanOriginationDate.year), color = I('black'),
      fill = I('#007EE5'), xlab = 'Year', ylab = 'Number of Loans') + 
      ggtitle('Number of Loans by Year') 
```
# Reflection

This is a complex dataset that I have been working on with a lot of variables and it's really tough to identify and imply correlation from it. Since, I am still a student and possibly hoping to get a job and take loans in the future, I am not aware of most of the terms in the dataset. Since the dataset was tidy I was able to perform a few calculation to arrive at the answers that I am looking for. This gives me a lot of confidence in the future if I am facing a challenge to perform EDA to an unknown dataset. The most important variable from this dataset according to my perspective was the lender yield, the debt to income ratio and prosper rating. I would still like to know more about how they rate the borrowers since there is an indifferent correlation with the lenders trust on the rating. I am looking at the website to find answer for this. The debt to income ratio plays an important role in the borrowers mindset to obtain loans and also the lenders yield. With this information, further linear regression and predictive modeling can be performed to arrive at a pattern for the lenders as well as borrowers to make the process of obtaining and loan and implication associated with it more easier.

#References
- Prosper: How it works: https://www.prosper.com/welcome/how-it-works/
- Prosper Ratings: https://www.prosper.com/invest/how-to-invest/prosper-ratings/
- What Is in a Credit Score?: https://www.creditkarma.com/article/credit-score-factors
- MORE DESCRIPTIVE STATISTICS:http://ww2.coastal.edu/kingw/statistics/R-tutorials/descriptive.html
- Beautiful plotting in R: A ggplot2 cheatsheet: http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/

